---
title: "Data Dowload"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

# OCLE physical variables function use

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
suppressPackageStartupMessages({
  invisible({
    library(ncdf4)
    library(raster)
    library(lubridate)
    library(tidyverse)
  })
})
```

# Download function

The data to be downloaded here is available thorugh a user interface
here: <https://ocle.ihcantabria.com/>

This tool helps you download climate data (like ocean temperatures, wind
speeds, etc.) of different types: - Past measurements ("Historical"
data) - Future projections ("Projected" data)

## Variables available

Historical options:

-   pH

-   Wind

-   TidalRange

-   Salinity

-   SST (sea surface temperature)

-   PAR (light availability)

-   Nitrate

-   MHW (marine heat waves)

-   MCS (marine cold spells)

-   Hs (wave heights)

-   Currents (ocean currents)

-   AirTemperature

Projected options (future):

-   pH

-   Wind

-   Salinity

-   SST

-   SLR (sea level rise)

-   Nitrates

-   Hs

-   AirTemperature

Output options:

-   ascii

-   nc

-   csv

-   rdata

Basic Requirements:

-   R software installed (free from <https://www.r-project.org/>)

-   These R packages installed: ncdf4, raster, lubridate, dplyr

-   Internet connection

```{r, echo=FALSE}
  
  download_data <- function(period, variable, parameter,
                            temporal_resolution ,
                            save_path, 
                            year_or_season = NULL,
                            scenario = NULL,
                            output_format = "nc",  # Options: "ascii", "nc", "csv", "rdata"
                            return_object = TRUE) {
    
    #____________________
    # Validate parameters
    valid_parameters <- c("Max", "Mean", "Min", "Percentile10", "Percentile50", "Percentile90")
    #____________________
    
    #_________________________
    # Validate output format
    #_________________________
    if (!output_format %in% c("ascii", "nc", "csv", "rdata")) {
      stop("Invalid output_format. Choose 'ascii', 'nc', 'csv', or 'rdata'")
    }
    
    # Projected variables restriction
    projected_vars <- c("pH", "Wind", "Salinity", "SST", "SLR", "Nitrates", "Hs", "AirTemperature")
    
    # Construct the base URL
    base_url <- "https://ihthredds.ihcantabria.com/thredds/dodsC/SIMULIT"
    
    #____________________________________________________
    # Validate period and variables
    #____________________________________________________
    if (period == "Historical") {
      available_vars <- c("pH", "Wind", "TidalRange", "ShearStress", "Salinity", "SST", "PAR", "Nitrate", 
                          "MHW", "MCS", "Hs", "Currents", "BottomOrbitalSpeed", "AttenuationCoefficient", "AirTemperature")
      if (!variable %in% available_vars) {
        stop(paste("Invalid variable for Historical period. Available options:", 
                   paste(available_vars, collapse = ", ")))
      }
      if (temporal_resolution %in% c("yearly", "winter", "summer", "spring", "autumn", "all")) {
        dataset_url <- paste0(base_url, "/", period, "/", variable, "/", variable, "_", temporal_resolution, ".nc")
      } else {
        stop("Invalid temporal resolution for Historical. Options: yearly, winter, summer, spring, autumn, all")
      }
    } else if (period == "Projected") {
      if (!variable %in% projected_vars) {
        stop(paste("Invalid variable for Projected period. Available options:", 
                   paste(projected_vars, collapse = ", ")))
      }
      if (scenario %in% c("SSP245", "SSP585") && year_or_season %in% c(2050, 2100)) {
        # Special URL structure for projected variables
        dataset_url <- paste0(base_url, "/", period, "/", variable, "/", scenario, "/", variable, ".nc")
      } else {
        stop("Invalid scenario or year for Projected. Scenario options: SSP245, SSP585. Year options: 2050, 2100")
      }
    } else {
      stop("Invalid period specified. Choose 'Historical' or 'Projected'")
    }
    
    if(temporal_resolution != "yearly"){
      warning("Temporal resolution ignored for this resolution.")
    }
    #_________________________
    # Opening the nc we want 
    nc <- nc_open(dataset_url)
    #_________________________
    
    #____________________________________________________
    #Dealing with vairables with no parameters to select 
    #____________________________________________________
    
    if (variable == "MHW" |variable == "MCS") {
      if(parameter %in% c("Percentile10", "Percentile50", "Percentile90")){
        warning("Parameter not available, select between: Max, Min or Mean")
        
      }
      # Warn about yearly-only availability
      if (temporal_resolution != "yearly") {
        warning("No parameter selection allowed. \nMHW data is only available yearly. Using yearly resolution instead of ", temporal_resolution)
      }
      var_name <- "MHW"  # Use direct variable name
    } 
    else if (variable == "SLR") {
      var_name <- "SLR"  # Use direct variable name
    }
    else {
      var_name <- paste0(variable, "_", parameter)  # Standard naming
    }
    
    #_________________________________________
    # Check if variable exists
    #_________________________________________
    
    if (!var_name %in% names(nc$var)) {
      available_vars <- names(nc$var)
      # Try to find closest match
      if (variable %in% available_vars) {
        var_name <- variable
        message("Using direct variable name '", variable, "' instead of '", 
                variable, "_", parameter, "'")
      } else {
        stop("Variable ", var_name, " not found. Available: ", paste(available_vars, collapse = ", "))
      }
    }
    #_________________________________________
    # Handle time dimension for projected data
    #_________________________________________
    
    if (period == "Projected") {
      time <- ncvar_get(nc, "time")
      time_vals <- as.POSIXct(time, origin = "1970-01-01")
      years <- as.numeric(format(time_vals, "%Y"))
      
      time_index <- which(years == year_or_season)
      if (length(time_index) == 0) {
        stop(paste("Year", year_or_season, "not found. Available years:", paste(unique(years), collapse = ", ")))
      }
      # filtering to keep the reuired year
      data <- ncvar_get(nc, var_name, start = c(1, 1, time_index), count = c(-1, -1, 1))
    } else {
      
      if(temporal_resolution == "yearly"){
        time <- ncvar_get(nc, "time")
        time_vals <- lubridate::as_datetime(time, tz = "UTC") #automatically sets teh 1970 start date
        year_required <- lubridate::year(time_vals)
        year_index <- which(year_required == year_or_season)
        request_var <- var_name #paste0(var_name, "_", parameter)
        
         # filtering to keep the reuired year
        data <- ncvar_get(nc, request_var, start = c(1, 1, year_index[1]), count = c(-1, -1, length(year_index)))
        
      }else{
        data <- ncvar_get(nc, var_name)
      }
    }
    
    #____________________
    # Get coordinates
    #____________________
    lon <- ncvar_get(nc, "lon")
    lat <- ncvar_get(nc, "lat")
    nc_close(nc)
    
    #______________________________________________
    # Creating the raster with the correct rotation 
    #_______________________________________________
    if (period == "Projected") { #Create raster with 90-degree counter-clockwise rotation
      rotated_matrix <- t(data)[ncol(data):1,]  # Rotate 90 degrees left
      r <- raster(rotated_matrix,
                  xmn = min(lon), xmx = max(lon),
                  ymn = min(lat), ymx = max(lat),
                  crs = CRS("+proj=longlat +datum=WGS84"))
    }
    if (period == "Historical" & temporal_resolution != "yearly") {
      rotated_matrix <- t(data)[,nrow(data):1]  # Simple right rotation
      r <- raster(rotated_matrix,
                  xmn = min(lon), xmx = max(lon),
                  ymn = min(lat), ymx = max(lat),
                  crs = CRS("+proj=longlat +datum=WGS84"))
      r <-  flip(r, direction = 'x')
    }
    
    if (period == "Historical" & temporal_resolution == "yearly") {
      # Rotate the matrix and create the raster
      rotated_matrix <- t(data)[ncol(data):1,]  # Rotate 90 degrees left
      r <- raster(rotated_matrix,
                  xmn = min(lon), xmx = max(lon),
                  ymn = min(lat), ymx = max(lat),
                  crs = CRS("+proj=longlat +datum=WGS84"))
    }
    
    # Create filename
    fname_base <- paste0(
      variable, 
      if (!variable %in% c("MHW", "MCS", "SLR")) paste0("_", parameter) else "",
      "_",
      if (period == "Historical" && temporal_resolution == "yearly") "yearly_" else "",
      year_or_season,
      if (period == "Projected") paste0("_", scenario) else ""
    )
    
    #__________________________________________________________________________
    # Checking if the directory given is actally there, if its not then make it
    #__________________________________________________________________________
    if (!dir.exists(save_path)) dir.create(save_path, recursive = TRUE)
    
    #_________________________________________
    # Output handling based on user selection
    #_________________________________________
    if (output_format == "ascii") {
      raster::writeRaster(r, file.path(save_path, paste0(fname_base, ".asc")), format = "ascii", overwrite = TRUE)
    } else if (output_format == "nc") {
      raster::writeRaster(r, file.path(save_path, paste0(fname_base, ".nc")), format = "CDF", overwrite = TRUE)
    } else {
      df <- as.data.frame(r, xy = TRUE)
      colnames(df) <- c("lon", "lat", "value")
      
      if (output_format == "csv") {
        write.csv(df, file.path(save_path, paste0(fname_base, ".csv")), row.names = FALSE)
      } else if (output_format == "rdata") {
        save(df, file = file.path(save_path, paste0(fname_base, ".RData")))
      }
    }
    
    if (return_object) {
      if (output_format %in% c("ascii", "nc")) return(r) else return(df)
    } else {
      return(invisible(NULL))
    }
    
  }
  
```

## 1. Setting up your workspace

First, set where you want to save files. Replace "your_folder_path" with
your actual folder path:

```{r set_path}
# Change this to your desired folder:
save_folder <- file.path(getwd(), "climate_data") 

# Create folder if it doesn't exist
if (!dir.exists(save_folder)) {
  dir.create(save_folder)
}

```

```{r set_path_output, echo=FALSE}
# Example output that would appear
cat("Your files will save to: /Users/yourname/Documents/climate_data")
```

------------------------------------------------------------------------

## 2. Downloading historical yearly data

Example: Get yearly maximum sea surface temperature (SST) for 2010

```{r historical_yearly}
result_sst <- download_data(
  period = "Historical",
  variable = "SST",
  parameter = "Max",
  temporal_resolution = "yearly",
  year_or_season = 2010,
  save_path = save_folder,
  output_format = "nc"
)

glimpse(result_sst)
```

*File saved at:* `climate_data/SST_Max_yearly_2010.nc`

------------------------------------------------------------------------

## 3. Downloading seasonal data

Example: Get summer mean pH values

```{r historical_seasonal}
result_ph <- download_data(
  period = "Historical",
  variable = "pH",
  parameter = "Mean",
  temporal_resolution = "summer",
  save_path = save_folder,
  output_format = "csv"
)

```

```{r echo =FALSE}

head(drop_na(result_ph), 3)
```

*File saved at:* `climate_data/pH_Mean_summer.csv`

------------------------------------------------------------------------

## 4. Downloading projected future data

Example: Get 2050 wind speed projections (SSP2 - 4.5 scenario)

```{r projected_data}
result_wind <- download_data(
  period = "Projected",
  variable = "Wind",
  parameter = "Mean",
  temporal_resolution = "yearly",
  year_or_season = 2050,
  scenario = "SSP245",
  save_path = save_folder,
  output_format = "rdata"
)

```

```{r echo =FALSE}

head(drop_na(result_wind), 3)
```

*File saved at:* `climate_data/Wind_Mean_yearly_2050_SSP245.RData`

------------------------------------------------------------------------

## 5. Marine Heat Waves (MHW)

Example: Get yearly marine heat wave data

```{r special_case}
result_mhw <- download_data(
  period = "Historical",
  variable = "MHW",
  parameter = "Mean",
  temporal_resolution = "yearly",
  year_or_season = 2015,
  save_path = save_folder,
  output_format = "ascii"
)

glimpse(result_mhw)
```

*File saved at:* `climate_data/MHW_Mean_yearly_2015.asc`

------------------------------------------------------------------------

## 6. Troubleshooting

Common errors:

-   "Variable not found": Check spelling and period compatibility

-   "Year not available": Historical data only goes to 2020

-   "Invalid parameter": Some variables have restrictions

-   Folder permissions: Make sure you can write to save_path
