---
title: "Data Dowload"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

# OCLE physical variables function use

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
suppressPackageStartupMessages({
  invisible({
    library(ncdf4)
    library(raster)
    library(lubridate)
    library(tidyverse)
  })
})
```

# Download function

The data to be downloaded here is available thorugh a user interface
here: <https://ocle.ihcantabria.com/>

This tool helps you download climate data (like ocean temperatures, wind
speeds, etc.) of different types: - Past measurements ("Historical"
data) - Future projections ("Projected" data)

## Variables available

Historical options:

-   pH

-   Wind

-   TidalRange

-   Salinity

-   SST (sea surface temperature)

-   PAR (light availability)

-   Nitrate

-   MHW (marine heat waves)

-   MCS (marine cold spells)

-   Hs (wave heights)

-   Currents (ocean currents)

-   AirTemperature

Projected options (future):

-   pH

-   Wind

-   Salinity

-   SST

-   SLR (sea level rise)

-   Nitrates

-   Hs

-   AirTemperature

Output options:

-   ascii

-   nc

-   csv

-   rdata

Basic Requirements:

-   R software installed (free from <https://www.r-project.org/>)

-   These R packages installed: ncdf4, raster, lubridate, dplyr

-   Internet connection

```{r, echo=FALSE}
 # Predefined available variables for each period
available_variables_historical <- c("pH", "Wind", "TidalRange", "ShearStress", "Salinity", "SST", "PAR", "Nitrate", "MHW", "MCS", "Hs", "Currents", "BottomOrbitalSpeed", "AttenuationCoefficient", "AirTemperature")
available_variables_projected <- c("pH", "Wind", "Salinity", "SST", "SLR", "Nitrates", "Hs", "AirTemperature")

# Function to get available variables based on the selected period
get_available_variables <- function(period) {
  if (period == "Historical") {
    return(available_variables_historical)
  } else if (period == "Projected") {
    return(available_variables_projected)
  } else {
    stop("Invalid period")
  }
}

download_data <- function(period, variable, parameter,
                         temporal_resolution, save_path,
                         year_or_season = NULL,
                         scenario = NULL,
                         lat_max = NULL,
                         lat_min = NULL,
                         lon_min = NULL,
                         lon_max = NULL,
                         output_format = "nc",  # Options: "ascii", "nc", "csv", "rdata"
                         return_object = TRUE
                         ) {
  
  # Load required packages
  if (!requireNamespace("ncdf4", quietly = TRUE)) install.packages("ncdf4")
  if (!requireNamespace("httr", quietly = TRUE)) install.packages("httr")
  if (!requireNamespace("raster", quietly = TRUE)) install.packages("raster")
  library(ncdf4)
  library(httr)
  library(raster)
  
  # Ensure save path exists
  if (!dir.exists(save_path)) dir.create(save_path, recursive = TRUE)
  
  #_________________________
  # Validate output format
  #_________________________
  if (!output_format %in% c("ascii", "nc", "csv", "rdata")) {
   stop("Invalid output_format. Choose 'ascii', 'nc', 'csv', or 'rdata'")
  }
 
  # Construct URL
  base_url <- "https://ihthredds.ihcantabria.com/thredds/fileServer/SIMULIT"
  
  if (period == "Historical") {
    dataset_url <- paste0(base_url, "/",
                          period, "/",
                          variable, "/",
                          variable, "_",
                          ifelse(temporal_resolution == "range", "yearly",
                                 ifelse(temporal_resolution == "full", "all", temporal_resolution)), 
                          ".nc")
  } else if (period == "Projected") {
    dataset_url <- paste0(base_url, "/",
                          period, "/",
                          variable, "/",
                          scenario, "/",#                          year_or_season,
                          variable,# "_",
                          ".nc")
  } else {
    stop("Invalid period")
  }
  
  #_________________________
  # Download file
  #_________________________
  temp_file <- tempfile(fileext = ".nc")
  response <- GET(
  dataset_url,
  write_disk(temp_file, overwrite = TRUE),
  config(followlocation = TRUE),  # Follow redirects (common in THREDDS)
  timeout(3000000),                   # Increase timeout (default: 60s)
  progress()
  )
  if (http_error(response)) stop("Failed to download file")
  
  # Open file
  nc <- nc_open(temp_file)
  on.exit({
    nc_close(nc)
    unlink(temp_file)
  })
  
  #_________________________
  # Get parameter name 
  #_________________________
  
  if(variable =="MHW"|variable =="MCS"){
    parameter <- variable
  full_param_name <- variable
   
  }else{
  full_param_name <- paste0(variable, "_", parameter)
      if (!full_param_name %in% names(nc$var)) {
    stop(paste("Parameter not found. Available parameters:",
               paste(names(nc$var), collapse = ", ")))
  }
  }
  

  
  #______________________________________________________________
  # Handle time dimension differently for seasonal vs yearly data
  #_________________________________________________________
  if (temporal_resolution == "yearly"|temporal_resolution == "range") {

    time_vals <- ncvar_get(nc, "time")
    time_units <- ncatt_get(nc, "time", "units")$value

    if (grepl("days since", time_units)) {
      time_dates <- as.Date(sub(".*since ", "", time_units)) + time_vals
    } else if (grepl("seconds since", time_units)) {
      time_dates <- as.POSIXct(sub(".*since ", "", time_units), tz="UTC") + as.difftime(time_vals, units="secs")
    } else {
      stop("Unrecognized time format")
    }

    time_years <- as.integer(format(time_dates, "%Y"))
    print(paste("Available years:", paste(unique(time_years), collapse=", ")))

    #______________________________
    # Selecting the required years 
    #______________________________
    if (length(year_or_season)== 1) {
      year_index <- which(time_years == as.numeric(year_or_season))
      if (length(year_index) == 0) stop("Requested year not available")
      data <- ncvar_get(nc, full_param_name,
                        start = c(1, 1, year_index),
                        count = c(-1, -1, 1))
    }else if(length(year_or_season) >1){
     # For multiple years
      year_indices <- which(time_years %in% year_or_season)
      
      # Get all the data for those years
      data <- ncvar_get(nc, full_param_name, 
                       start = c(1, 1, min(year_indices)), 
                       count = c(-1, -1, length(year_indices)))
            
    } else {
      data <- ncvar_get(nc, full_param_name)
    }
    
  }else {
    # Seasonal data - no time dimension to process
    data <- ncvar_get(nc, full_param_name)
  }
   

  #______________________________________________________
  # Create output file
  #______________________________________________________
  lon <- ncvar_get(nc, "lon")
  lat <- ncvar_get(nc, "lat")
  
  #________________________
  # Filtering by dimensions 
  #________________________
  if (!is.null(lat_max) || !is.null(lat_min) || !is.null(lon_min) || !is.null(lon_max)) {
    
    # Set defaults if NULL (entire range)
    if (is.null(lat_max)) lat_max <- max(lat)
    if (is.null(lat_min)) lat_min <- min(lat)
    if (is.null(lon_max)) lon_max <- max(lon)
    if (is.null(lon_min)) lon_min <- min(lon)
    
    # Find indices for subsetting
    lat_indices <- which(lat >= lat_min & lat <= lat_max)
    lon_indices <- which(lon >= lon_min & lon <= lon_max)
    lon <- lon[lon_indices]
    lat <- lat[lat_indices]
    
  if(period == "Projected"){
    data <- data[lon_indices, lat_indices,]  
  }else{
        # Because year adds another dimension
    if (length(year_or_season)== 1) {
      data <- data[lon_indices, lat_indices]  
    }else if(length(year_or_season) >1){
      data <- data[lon_indices, lat_indices,]  
    }
    
  }

    
  }
  
  # Create raster object
  rotated_matrix <- t(data)[ncol(data):1,]  # Rotate 90 degrees left
  r <- raster(rotated_matrix,
              xmn = min(lon), xmx = max(lon),
              ymn = min(lat), ymx = max(lat),
              crs = CRS("+proj=longlat +datum=WGS84"))
  
  # Create filename
  fname_base <- paste0(
    variable, "_", parameter, "_", period, "_", temporal_resolution,
    ifelse(!is.null(year_or_season), 
           ifelse(temporal_resolution == "range", 
                  paste0("_", min(year_or_season), "_", max(year_or_season)),
                  paste0("_", year_or_season)),
           ""),
    ifelse(!is.null(scenario), paste0("_", scenario), "")
  )
  
  #_________________________________________
  # Output handling based on user selection
  #_________________________________________
  if (output_format == "ascii") {
    raster::writeRaster(r, file.path(save_path, paste0(fname_base, ".asc")), format = "ascii", overwrite = TRUE)
  } else if (output_format == "nc") {
    raster::writeRaster(r, file.path(save_path, paste0(fname_base, ".nc")), format = "CDF", overwrite = TRUE)
  } else {
    df <- as.data.frame(r, xy = TRUE)
    colnames(df) <- c("lon", "lat", "value")
    
    if (output_format == "csv") {
      write.csv(df, file.path(save_path, paste0(fname_base, ".csv")), row.names = FALSE)
    } else if (output_format == "rdata") {
      save(df, file = file.path(save_path, paste0(fname_base, ".RData")))
    }
  }
  
  if (return_object) {
    if (output_format %in% c("ascii", "nc")) return(r) else return(df)
  } else {
    return(invisible(NULL))
  }
}
```

## 1. Setting up your workspace

First, set where you want to save files. Replace "your_folder_path" with
your actual folder path:

```{r set_path}
# Change this to your desired folder:
save_folder <- file.path(getwd(), "climate_data") 

# Create folder if it doesn't exist
if (!dir.exists(save_folder)) {
  dir.create(save_folder)
}

```

```{r set_path_output, echo=FALSE}
# Example output that would appear
cat("Your files will save to: /Users/yourname/Documents/climate_data")
```

------------------------------------------------------------------------

## 2. Downloading historical yearly data

Example: Get yearly maximum sea surface temperature (SST) for 2010

```{r historical_yearly}
result_sst <- download_data(
  period = "Historical",
  variable = "SST",
  parameter = "Max",
  temporal_resolution = "yearly",
  year_or_season = 2010,
  save_path = save_folder,
  output_format = "nc"
)

glimpse(result_sst)
```

*File saved at:* `climate_data/SST_Max_yearly_2010.nc`

------------------------------------------------------------------------

## 3. Downloading seasonal data

Example: Get summer mean pH values

```{r historical_seasonal}
result_ph <- download_data(
  period = "Historical",
  variable = "pH",
  parameter = "Mean",
  temporal_resolution = "summer",
  save_path = save_folder,
  output_format = "csv"
)

```

```{r echo =FALSE}

head(drop_na(result_ph), 3)
```

*File saved at:* `climate_data/pH_Mean_summer.csv`

------------------------------------------------------------------------

------------------------------------------------------------------------

## 4. Downloading full data

Example: Get summer mean pH values

```{r historical_full}
result_pH_full <- download_data(
  period = "Historical",
  variable = "pH",
  parameter = "Max",
  temporal_resolution = "full",
  save_path = save_folder,
  output_format = "csv"  # Now can also be "ascii", "csv", or "rdata"
)

```

```{r echo =FALSE}

head(drop_na(result_pH_full), 3)
```

*File saved at:* `climate_data/pH_Mean_summer.csv`

------------------------------------------------------------------------
## 5. Downloading full data for a region

Example: Get summer mean temperature for the whole time period for a particular area.

```{r historical_full_galicia}

# Example for Galicia coast (NW Spain)
galicia_result <- download_data(
  period = "Historical",
  variable = "SST",               # Sea Surface Temperature
  parameter = "Mean",             # We want the mean values
  temporal_resolution = "full",   # Full time series
  save_path = "~/climate_data",   # Your save path
  lat_min = 41.5,                 # Southern boundary (approx. Ría de Vigo)
  lat_max = 44.0,                 # Northern boundary (approx. A Coruña)
  lon_min = -9.5,                 # Western boundary
  lon_max = -7.5,                 # Eastern boundary
  output_format = "nc"            # Can change to "csv", "rdata", etc.
)

glimpse(galicia_result)

```

------------------------------------------------------------------------

## 6. Downloading projected future data

Example: Get 2050 wind speed projections (SSP2 - 4.5 scenario)

```{r projected_data}
result_wind <- download_data(
  period = "Projected",
  variable = "Wind",
  parameter = "Mean",
  temporal_resolution = "yearly", #its for the period of the year selected
  year_or_season = 2050,
  scenario = "SSP245",
  save_path = save_folder,
  output_format = "rdata"
)

```

```{r echo =FALSE}

head(drop_na(result_wind), 3)
```

*File saved at:* `climate_data/Wind_Mean_yearly_2050_SSP245.RData`

------------------------------------------------------------------------

## 7. Marine Heat Waves (MHW)

Example: Get yearly marine heat wave data

```{r special_case}
result_mhw <- download_data(
  period = "Historical",
  variable = "MHW",
  parameter = "Mean",
  temporal_resolution = "yearly",
  year_or_season = 2015,
  save_path = save_folder,
  output_format = "ascii"
)

glimpse(result_mhw)
```

*File saved at:* `climate_data/MHW_Mean_yearly_2015.asc`

------------------------------------------------------------------------

## 8. Troubleshooting

Common errors:

-   "Variable not found": Check spelling and period compatibility

-   "Year not available": Historical data only goes to 2020

-   "Invalid parameter": Some variables have restrictions

-   Folder permissions: Make sure you can write to save_path
